---
title: "Benchmarking Tensory TTM Functions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Benchmarking Tensory TTM Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Benchmarking Approach

This vignette describes the benchmarking approach used to compare the performance of two implementations of the tensor-times-matrix operation in the tensory package:

- `tensory:::ttm_test_cpp` (original implementation)
- `tensory:::ttm_test1_cpp` (optimized implementation)

## Benchmark Components

### Basic Performance Comparison

We start with a basic performance comparison using `bench::mark` with a fixed-size example based on the user's provided example:

```{r basic-example, eval=FALSE}
# Create a random 3-mode tensor and a matrix based on user's example
X <- array(runif(30*40*50), dim = c(30, 40, 50))
mat <- matrix(runif(20*40), nrow = 20, ncol = 40)

# Prepare tensory tensor object
X_tensory <- tensory::Tensor$new(X)

# Benchmark the functions
bm <- bench::mark(
  ttm_test_cpp = tensory:::ttm_test_cpp(X_tensory$data, mat, mode = 2),
  ttm_test1_cpp = tensory:::ttm_test1_cpp(X_tensory$data, mat, mode = 2),
  min_iterations = 100,
  check = FALSE
)
```

### Parameterized Benchmark

We use `bench::press` to test performance across different tensor dimensions:

```{r parameterized-example, eval=FALSE}
# Define parameter grid
params <- expand.grid(
  dim1 = c(20, 40, 60),
  dim2 = c(30, 50, 70),
  dim3 = c(25, 45, 65),
  mat_rows = c(15, 35, 55),
  #mat_cols = c(30, 50, 70),
  mode = c(1, 2, 3),
  stringsAsFactors = FALSE
)

# Filter to ensure matrix columns match tensor dimension for the mode
valid_params <- params[params$mat_cols == ifelse(params$mode == 1, params$dim1, 
                                                 ifelse(params$mode == 2, params$dim2, params$dim3)), , drop = FALSE]

# Take a sample of parameter combinations to reduce computation time
if (nrow(valid_params) > 20) {
  valid_params <- valid_params[sample(nrow(valid_params), 20), , drop = FALSE]
}

# Run parameterized benchmark
bp <- bench::press(
  dim1 = valid_params$dim1,
  dim2 = valid_params$dim2,
  dim3 = valid_params$dim3,
  mat_rows = valid_params$mat_rows,
  mat_cols = valid_params$mat_cols,
  #mode = valid_params$mode,
  {
    # Create tensor and matrix based on parameters
    X <- array(runif(dim1 * dim2 * dim3), dim = c(dim1, dim2, dim3))
    mat <- matrix(runif(mat_rows * mat_cols), nrow = mat_rows, ncol = mat_cols)
    
    # Prepare tensory tensor object
    #X_tensory <- tensory::Tensor$new(X)
    
    # Benchmark both functions
    bench::mark(
      ttm_test_cpp = tensory:::ttm_test_cpp(X, mat, mode = mode),
      ttm_test1_cpp = tensory:::ttm_test1_cpp(X, mat, mode = mode),
      min_iterations = 10,
      check = FALSE
    )
  }
)
```

### Transpose Operations Benchmark

We also evaluate performance with and without transpose operations:

```{r transpose-example, eval=FALSE}
# Create test data
X <- array(runif(25*35*45), dim = c(25, 35, 45))
mat <- matrix(runif(30*35), nrow = 30, ncol = 35)  # For transpose = FALSE
mat_t <- matrix(runif(35*30), nrow = 35, ncol = 30)  # For transpose = TRUE

# Prepare tensory tensor object
#X_tensory <- tensory::Tensor$new(X)

# Benchmark with transpose variations
bt <- bench::press(
  transpose = c(TRUE, FALSE),
  {
    if (transpose) {
      bench::mark(
        ttm_test_cpp = tensory:::ttm_test_cpp(X, mat_t, mode = 2, transpose = TRUE),
        ttm_test1_cpp = tensory:::ttm_test1_cpp(X, mat_t, mode = 2, transpose = TRUE),
        iterations = 5000,
        check = FALSE
      )
    } else {
      bench::mark(
        ttm_test_cpp = tensory:::ttm_test_cpp(X, mat, mode = 2, transpose = FALSE),
        ttm_test1_cpp = tensory:::ttm_test1_cpp(X, mat, mode = 2, transpose = FALSE),
        iterations = 5000,
        check = FALSE
      )
    }
  }
)
```

### Memory Usage Analysis

We analyze memory allocation differences between implementations:

```{r memory-example, eval=FALSE}
# Create test data
X <- array(runif(20*30*40), dim = c(20, 30, 40))
mat <- matrix(runif(25*30), nrow = 25, ncol = 30)

# Prepare tensory tensor object
X_tensory <- tensory::Tensor$new(X)

# Benchmark with memory tracking
bm_memory <- bench::mark(
  ttm_test_cpp = tensory:::ttm_test_cpp(X_tensory$data, mat, mode = 2),
  ttm_test1_cpp = tensory:::ttm_test1_cpp(X_tensory$data, mat, mode = 2),
  min_iterations = 50,
  memory = TRUE,
  check = FALSE
)
```

## Running the Benchmarks

To run the full benchmark suite, use the provided scripts in the `benchmarks/` directory:

```{r run-benchmarks, eval=FALSE}
source("benchmarks/ttm_benchmark.R")
results <- run_all_benchmarks()
```

## Results

The benchmark results are saved in the `benchmark_results/` directory, including:

- CSV files with raw benchmark data
- PNG visualizations of performance comparisons
- A markdown report summarizing the findings

## Conclusion

The `ttm_test1_cpp` implementation shows improved performance over the original `ttm_test_cpp` implementation, with optimizations including:

1. Avoiding forced column-major layout conversion
2. Using auto type deduction to prevent unnecessary layout conversions
3. More efficient memory access patterns
4. Reduced memory allocation overhead

These optimizations result in significant speedup in typical use cases.