---
title: "performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Computational Efficiency: ttm vs rTensor

We compare the computational efficiency of the `ttm` function from the `tensory` package with the equivalent function in the `rTensor` package. This benchmark demonstrates the performance benefits of using `tensory` for tensor-times-matrix multiplication.

```{r efficiency-comparison, message=FALSE, warning=FALSE}
set.seed(123)

# Create a random 3-mode tensor and a matrix
X <- array(runif(50*30*50), dim = c(50, 30, 50))
mat <- matrix(runif(70*30), nrow = 70, ncol = 30)

library(bench)

# Prepare rTensor object
X_rTensor <- rTensor::as.tensor(X)
X <- tensory::Tensor$new(X)

# Benchmark ttm in tensory vs rTensor
bm <- bench::mark(
  tensory = tensory::ttm(X, mat, mode = 2),
  rTensor = rTensor::ttm(X_rTensor, mat, m = 2),
  iterations = 20,
  check = FALSE
)
bm

# Visualization
library(ggplot2)
autoplot(bm)
```

## Multiple Matrix Multiplications Benchmark

Here we benchmark the performance of multiple matrix multiplications, which is a common operation in tensor decomposition algorithms and multilinear algebra.

```{r multiple-matrix-benchmark, message=FALSE, warning=FALSE}
set.seed(456)

# Create a 4-mode tensor and matrices for each mode
X <- array(runif(50*40*30*20), dim = c(50, 40, 30, 20))
matrices <- list(
  matrix(runif(60*50), nrow = 60, ncol = 50),  # mode 1
  matrix(runif(45*40), nrow = 45, ncol = 40),  # mode 2
  matrix(runif(35*30), nrow = 35, ncol = 30),  # mode 3
  matrix(runif(25*20), nrow = 25, ncol = 20)   # mode 4
)

library(bench)

# Prepare rTensor object
X_rTensor <- rTensor::as.tensor(X)
X <- tensory::Tensor$new(X)

# Define function for multiple rTensor ttm operations
multiple_ttm_rTensor <- function(tensor, mat_list) {
  result <- tensor
  for (i in seq_along(mat_list)) {
    result <- rTensor::ttm(result, mat_list[[i]], m = i)
  }
  return(result)
}

# Benchmark multiple ttm operations
bm_multiple <- bench::mark(
  tensory = tensory::ttm(X, matrices, mode = c(1, 2, 3, 4)),
  rTensor = multiple_ttm_rTensor(X_rTensor, matrices),
  iterations = 10,
  check = FALSE
)
bm_multiple

# Visualization
library(ggplot2)
autoplot(bm_multiple)
```

## Memory Efficiency Comparison

Let's also compare memory usage during these operations:

```{r memory-benchmark, message=FALSE, warning=FALSE}
set.seed(789)

# Create a tensor and matrix for memory comparison
X <- array(runif(30*25*20), dim = c(30, 25, 20))
mat <- matrix(runif(35*25), nrow = 35, ncol = 25)

library(bench)

# Prepare rTensor object
X_rTensor <- rTensor::as.tensor(X)
X <- tensory::Tensor$new(X)

# Memory benchmark
bm_memory <- bench::mark(
  tensory = tensory::ttm(X, mat, mode = 2),
  rTensor = rTensor::ttm(X_rTensor, mat, m = 2),
  iterations = 50,
  check = FALSE
)
bm_memory

# Visualization
library(ggplot2)
autoplot(bm_memory)
```

## Scalability Analysis

Finally, let's analyze how performance scales with tensor size:

```{r scalability-benchmark, message=FALSE, warning=FALSE}
set.seed(101112)

# Test different tensor sizes
sizes <- c(20, 30, 40, 50)
scalability_results <- data.frame()

library(bench)

for (size in sizes) {
  # Create tensor of varying size
  X <- array(runif(size^3), dim = c(size, size, size))
  mat <- matrix(runif((size+10) * size), nrow = size+10, ncol = size)
  
  # Prepare rTensor object
  X_rTensor <- rTensor::as.tensor(X)
  X <- tensory::Tensor$new(X)
  
  # Benchmark
  bm_scale <- bench::mark(
    tensory = tensory::ttm(X, mat, mode = 2),
    rTensor = rTensor::ttm(X_rTensor, mat, m = 2),
    iterations = 10,
    check = FALSE
  )
  
  # Store results
  temp_results <- data.frame(
    size = size,
    package = as.character(bm_scale$expression),
    median_time = as.numeric(bm_scale$median),
    memory = as.numeric(bm_scale$mem_alloc)
  )
  
  scalability_results <- rbind(scalability_results, temp_results)
}

# Visualization
library(ggplot2)
ggplot(scalability_results, aes(x = size, y = median_time, color = package)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  scale_y_log10() +
  labs(title = "Scalability Comparison",
       subtitle = "Performance vs Tensor Size",
       x = "Tensor Dimension Size",
       y = "Median Time (log scale)",
       color = "Package") +
  theme_minimal()
```
