---
title: "Quantile Regression with Tensor Predictors: pqtr()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantile Regression with Tensor Predictors: pqtr()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 3.2)
```

```{r setup}
library(tensory)
```

# The problem this solves

Each subject contributes a whole **array** — an image, a region-by-region
connectivity matrix, a sensor-by-time grid — and one continuous outcome. The
usual question is "how does the array move the *average* outcome?". `pqtr()`
asks a different one: **how does the array move a particular quantile of the
outcome?**

That distinction matters whenever the interesting subjects are not the average
ones. A brain region may barely shift mean cognitive score while strongly
predicting who ends up in the bottom decile. A predictor may leave the median
alone and widen the spread. Mean regression cannot see either effect; fitting
`tau = 0.1` and `tau = 0.9` and comparing the two coefficient arrays can.

Quantile regression is also robust: it makes no assumption about the error
distribution and is unmoved by outliers in the outcome, which is worth having
on its own when the response is skewed or heavy-tailed.

The obstacle is size. A 32 x 32 image is 1,024 predictors; with 200 subjects
the flattened quantile regression has no unique solution, and flattening also
throws away the grid — it forgets that row 4 column 12 sits next to row 4
column 13. `pqtr()` keeps the grid, compressing each *dimension* of the array
separately into a few directions chosen for their association with the
quantile of interest, then mapping the answer back to the original shape.

**Which function do I want?**

|                       | `pqtr()`                    | `tepls()`                  | `spgtr()`                       |
|-----------------------|-----------------------------|----------------------------|---------------------------------|
| Models                | a quantile of the outcome   | the mean                   | a GLM mean (any family)         |
| Outcome               | continuous                  | continuous, one or several | binary, count, continuous, ...  |
| Extra covariates      | yes, unreduced              | no                         | yes, unpenalized                |
| Selects whole slices  | no                          | no                         | yes, via an L2,1 penalty        |
| Robust to outliers    | yes                         | no                         | no                              |

# How to lay out your data

Two forms are accepted, and they are interchangeable.

**A list, one array per subject.** Every element must have the same
dimensions.

```{r}
set.seed(1)
p <- c(16, 12)
n <- 300
X <- lapply(seq_len(n), function(i) matrix(rnorm(prod(p)), p[1], p[2]))
length(X)
dim(X[[1]])
```

**One big array with subjects in the *last* mode.** If your data already
arrives as a `16 x 12 x 300` block, hand it over directly.

```{r}
Xbig <- array(unlist(X), dim = c(p, n))
dim(Xbig)
```

The outcome is a numeric vector of length `n`. Unlike `tepls()`, one outcome
at a time — a quantile is defined for a scalar.

# Quick start

We plant a signal in the top-left corner of the image, and make the noise grow
with it. The corner therefore shifts the median, and shifts the upper tail by
more: the coefficient array genuinely depends on which quantile you ask about.

```{r}
B_true <- outer(c(2, 1, rep(0, p[1] - 2)), c(1.5, rep(0, p[2] - 1)))
signal <- vapply(X, function(xi) sum(B_true * xi), numeric(1))
y <- signal + 0.4 * (1 + 0.8 * (signal - min(signal))) * rnorm(n)

med <- pqtr(X, y, tau = 0.5)
med
```

That is the whole call. `u` was not supplied, so `pqtr()` chose the number of
directions per mode itself. The coefficient is an array of the same shape as
one subject's data:

```{r}
B_med <- as.tensor(coef(med))$as_array()
dim(B_med)
round(B_med[1:4, 1:4], 2)
```

Now the same data at three quantiles. We fix `u` here so the three fits are
compared on equal terms:

```{r}
fits <- lapply(c(0.1, 0.5, 0.9), function(t) pqtr(X, y, tau = t, u = c(1, 1)))
names(fits) <- c("tau = 0.1", "tau = 0.5", "tau = 0.9")
vapply(fits, function(f) sqrt(sum(f$bvec^2)), numeric(1))
```

The coefficient grows as we move up the distribution, which is exactly how the
data were generated. Pictures make it plain:

```{r, fig.height = 2.6}
op <- par(mfrow = c(1, 4), mar = c(1, 1, 2, 1))
image(t(B_true[p[1]:1, ]), main = "truth", axes = FALSE)
for (nm in names(fits)) {
  image(t(as.tensor(coef(fits[[nm]]))$as_array()[p[1]:1, ]), main = nm,
        axes = FALSE)
}
par(op)
```

Predictions are fitted **conditional quantiles**, not conditional means:

```{r}
q10 <- predict(fits[["tau = 0.1"]])
q90 <- predict(fits[["tau = 0.9"]])
c(below_q10 = mean(y < q10), below_q90 = mean(y < q90))
```

About 10% of subjects fall below the fitted 10th percentile and about 90%
below the fitted 90th. That calibration is automatic — it is what minimizing
the check loss enforces — and it is a quick sanity check on any fit.

The gap between the two is a **prediction interval** for a new subject, one
that widens where the model says the outcome is more variable:

```{r}
width <- q90 - q10
summary(width)
```

# Choosing the number of directions

`u` is the one knob: how many directions to keep in each mode. `u = c(1, 1)`
keeps one row-pattern and one column-pattern; `u = c(3, 2)` keeps three and
two. Larger `u` means a more flexible fit, with `prod(u)` latent scores fed to
the quantile regression.

Left at its default `NULL`, `pqtr()` uses an eigenvalue-ratio rule: for each
mode it takes the largest gap between consecutive eigenvalues of that mode's
signal matrix, searching a few candidates.

```{r}
med$u
```

When you know the structure, say so — a scalar is recycled across modes:

```{r}
pqtr(X, y, tau = 0.5, u = c(1, 1))$u
pqtr(X, y, tau = 0.5, u = 2)$u
```

When you do not, cross-validate. `pqtr_cv()` scores a grid by out-of-fold
check loss and returns the model refitted at the winner:

```{r}
set.seed(2)
cv <- pqtr_cv(X, y, tau = 0.5, u_grid = 1:4, nfolds = 5)
cv$cv
cv$u_min
```

The default grid keeps the same number of directions in every mode. To let the
modes differ, pass the candidates yourself:

```{r}
grid <- list(c(1, 1), c(2, 1), c(1, 2), c(2, 2))
pqtr_cv(X, y, tau = 0.5, u_grid = grid, nfolds = 3)$u_min
```

Prefer the smallest `u` within noise of the best: extra directions cost
parameters in the quantile regression and start fitting noise.

# Covariates that should not be reduced

Age, sex, batch, scanner — variables you want adjusted for but not compressed
— go in `Z`. They enter the model unreduced, and they also enter the *first*
step, where the outcome is replaced by its working residual from a quantile
regression on `Z` alone. That is what makes the extracted directions describe
the part of the array not already explained by the covariates.

```{r}
Z <- cbind(age = rnorm(n), male = rbinom(n, 1, 0.5))
y_adj <- y + Z %*% c(2, -1)

fit_z <- pqtr(X, as.vector(y_adj), tau = 0.5, Z = Z, u = c(1, 1))
round(fit_z$gamma, 2)
```

`predict()` then needs the new covariates as well:

```{r}
head(predict(fit_z, X[1:5], Z[1:5, , drop = FALSE]))
```

# What is in the fit

```{r}
names(med)
```

* `coef` — the coefficient array as a `TTensor`, with core `core` and weight
  matrices `W`. `coef(fit)` returns it; `as.tensor(coef(fit))` expands it to a
  dense array; `fit$bvec` is the same thing flattened.
* `W` — one matrix per mode, `p_k x u[k]`, whose columns are the directions
  kept in that mode. These have orthonormal columns.
* `alpha`, `gamma` — intercept and covariate coefficients, on the centered
  scale (`predict()` handles the centering for you).
* `u`, `dims`, `tau`, `n` — the shape of the problem.
* `fitted`, `residuals`, `loss` — fitted conditional quantiles, their
  residuals, and the attained check loss.
* `scores` — the `n x prod(u)` compressed coordinates, useful for plotting or
  as input to another model.

The per-mode directions are interpretable on their own. Here the first mode's
direction should concentrate on rows 1 and 2 and the second mode's on column 1:

```{r}
round(med$W[[1]][1:4, 1, drop = FALSE], 3)
round(med$W[[2]][1:4, 1, drop = FALSE], 3)
```

# Evaluating honestly

In-sample calibration is guaranteed by construction, so it proves nothing.
Hold data out and score with the same check loss the method minimizes: for a
fitted quantile `q`, the loss on a new subject is `(y - q) * (tau - 1{y < q})`.

```{r}
set.seed(3)
tr <- sample(n, 200)
te <- setdiff(seq_len(n), tr)
check <- function(r, tau) mean(r * (tau - (r < 0)))

f <- pqtr(X[tr], y[tr], tau = 0.5, u = c(1, 1))
c(in_sample  = check(y[tr] - predict(f), 0.5),
  out_sample = check(y[te] - predict(f, X[te]), 0.5))
```

A useful baseline is the constant fit, which ignores the array entirely:

```{r}
check(y[te] - median(y[tr]), 0.5)
```

`predict()` accepts new data in either input form and checks the dimensions:

```{r}
identical(predict(f, X[te]),
          predict(f, array(unlist(X[te]), dim = c(p, length(te)))))
try(predict(f, lapply(1:5, function(i) matrix(0, 3, 3))))
```

# A caveat: quantile crossing

Each quantile is fit separately, so nothing forces the fitted 10th percentile
to stay below the fitted 90th for every subject. In practice crossings are
rare and confined to the edges of the predictor space, but they are worth
checking:

```{r}
mean(q90 < q10)
```

If crossings do appear, they usually mean `u` is too large for the sample size
at one of the quantiles, or that the extreme `tau` you asked for is not
supported by enough data.

# How it works

`pqtr()` implements the algorithm of Sun et al. (2024).

1. **Nuisance fit.** Fit the quantile regression of the outcome on `Z` alone
   (just the intercept when `Z` is absent, which is the sample quantile), and
   form the working residual `tau - 1{y below that fit}`. This is the
   subgradient of the check loss, and it is the only way the outcome enters
   the dimension-reduction step.
2. **Center** the predictor.
3. **Marginal covariances.** For each mode `k`, `Sigma_k` averages
   `X_(k) X_(k)'` over subjects — how variable the array is along that mode.
   (This step runs in C++ when the compiled kernel is available.)
4. **Partial quantile covariance.** `C` is the array of covariances between
   each predictor cell and the working residual — where the quantile signal
   is.
5. **Per-mode signal matrix.** `U_k = C_(k) (kron_{j != k} Sigma_j^-1) C_(k)'`
   standardizes `C` along every mode but `k`, so `U_k` measures signal in mode
   `k` after accounting for the others.
6. **Deflation.** The leading eigenvector of `U_k` is the first direction;
   project it out and repeat, `u[k]` times, giving the weight matrix `W_k`.
7. **Reduced quantile regression.** Project every subject onto
   `W_1 kron ... kron W_m` and fit the quantile regression of the outcome on
   the covariates and the resulting `prod(u)` scores.
8. **Map back.** The coefficient array is the latent coefficient expanded
   through the weight matrices, `B = D x_1 W_1 ... x_m W_m`.

Nothing here inverts a `prod(p) x prod(p)` matrix — only the per-mode
`p_k x p_k` ones — which is why the number of subjects can be far smaller than
the number of cells in the array. Only the dimension-reduction step is
approximate; the final fit is an exact linear quantile regression on the
scores.

**Relation to the reference implementation.** This is a port of the MATLAB
code released with the paper (<https://github.com/dayusun/PQTR>), with three
deliberate differences.

* The inner quantile regressions use the MM algorithm of Hunter & Lange
  (2000), whose surrogate is matched to the check loss, rather than MATLAB's
  `fminunc`, a smooth solver applied to a non-smooth objective.
* The predictor is centered before the latent scores are formed. This changes
  what `alpha` means (it is now the intercept at the training means) but
  leaves the coefficient array identical.
* Cross-validated selection lives in `pqtr_cv()`, which takes an explicit grid
  instead of enumerating every combination of per-mode dimensions — the
  default grid uses one common dimension, and anything else you can pass by
  hand.
* The eigenvalue-ratio rule searches at most five candidate dimensions per
  mode instead of about `sqrt(n)` of them. Past the rank of the signal matrix
  the eigenvalues are noise, and the largest ratio among them wins: on the
  data above the wider search returns `u = c(12, 11)`, which is 132 latent
  scores from 300 subjects. This is the same cap `spgtr()` and `tepls()` use.

The deflation in step 6 is the one used by the reference code: the oblique
projector `I - Sigma_k W (W' Sigma_k W)^-1 W'`, which leaves the weight
matrices with orthonormal columns. It is *not* the SIMPLS deflation used by
`tepls()` and `spgtr()`, which instead makes the latent scores uncorrelated.
The two coincide when the signal matrix has rank one and differ otherwise.

# Limits

* **One outcome at a time.** A quantile is defined for a scalar response; for
  several continuous outcomes at once, see `tepls()`.
* **No sparsity.** Every cell of the array contributes. For "which slices
  matter", use `spgtr()` with `lambda > 0` — at the cost of modelling the mean
  rather than a quantile.
* **No standard errors.** The coefficient array is a point estimate.
  Bootstrap the subjects if you need uncertainty.
* **Extreme quantiles need data.** A `tau` of 0.01 with 200 subjects rests on
  a couple of observations. Keep `tau` well inside the range the sample can
  support.

# References

Sun, D., Zhang, X. and Zhang, S. (2024). Partial quantile tensor regression.
*Journal of the American Statistical Association* **120**(551).

Hunter, D. R. and Lange, K. (2000). Quantile regression via an MM algorithm.
*Journal of Computational and Graphical Statistics* **9**(1), 60–77.
