---
title: "Tensor Regression with spgtr()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tensor Regression with spgtr()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 3.2)
```

```{r setup}
library(tensory)
```

# The problem this solves

Suppose every subject in your study contributes a whole **array** of numbers
rather than a handful of variables:

* an image, or a slice of one;
* a brain connectivity matrix (region by region);
* a spectrogram, or any sensor-by-time grid;
* a spatial grid measured at several time points.

and each subject also has one **outcome**: a diagnosis (yes/no), a count, a
score. You want to know how the array relates to the outcome, and which parts
of the array matter.

You cannot simply flatten the array and run `glm()`. A modest 60 × 60 image is
3,600 predictors; with 300 subjects, ordinary logistic regression has no unique
solution and any answer it produces is noise. Flattening also throws away the
grid structure: it forgets that column 12 of row 4 sits next to column 13 of
row 4.

`spgtr()` keeps the structure. It compresses each *dimension* of the array
separately down to a few informative directions, fits a generalized linear
model on the compressed version, and then translates the fitted model back to
the original array shape so you can look at it. Adding a sparsity penalty makes
it also *select*: whole rows, columns, or slices that carry no signal are
dropped, and you are told which ones survived.

# Data layout

Two input shapes are accepted, whichever is more convenient:

1. **A list**, `X[[i]]` being subject `i`'s matrix or array. All subjects must
   share the same shape.
2. **One big array** of order `m + 1`, with subjects in the **last**
   dimension, e.g. `dim(X) == c(8, 6, 120)` for 120 subjects measured on an
   8 × 6 grid.

The outcome `y` is a plain vector with one entry per subject: `0/1`, `TRUE`/
`FALSE`, or a two-level factor for yes/no outcomes; counts for `poisson()`;
any numbers for `gaussian()`.

Let us simulate 200 subjects measured on an 8 × 6 grid where only the top-left
corner drives a yes/no outcome.

```{r simulate}
set.seed(1)
n <- 200
p <- c(8, 6)

B_true <- matrix(0, p[1], p[2])
B_true[1, 1:2] <- c(2, 1.5) # only two cells carry signal

X <- lapply(seq_len(n), function(i) matrix(rnorm(prod(p)), p[1], p[2]))
eta <- vapply(X, function(xi) sum(B_true * xi), numeric(1))
y <- rbinom(n, 1, 1 / (1 + exp(-eta)))

length(X)
dim(X[[1]])
table(y)
```

# Quick start

One call fits the model. `u` is the number of directions kept per dimension;
leave it out and it is chosen for you.

```{r fit}
fit <- spgtr(X, y, u = c(1, 1))
fit
```

`summary()` adds a plain report of what was kept and how well the model fits.

```{r summary}
summary(fit)
```

The estimated coefficient array has exactly the shape of one subject's data.
Large entries are the cells that push the outcome up or down.

```{r coefs}
B_hat <- as.tensor(coef(fit))$as_array()
round(B_hat, 2)
```

```{r coefplot, fig.alt = "True and estimated coefficient grids side by side"}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
image(t(B_true[nrow(B_true):1, ]), main = "truth", axes = FALSE)
image(t(B_hat[nrow(B_hat):1, ]), main = "spgtr estimate", axes = FALSE)
par(op)
```

Predictions come in three flavours.

```{r predict}
head(predict(fit, type = "response")) # probabilities
head(predict(fit, type = "class")) # 0/1 labels
head(predict(fit, type = "link")) # log-odds
```

New subjects go in exactly like the training data:

```{r predict-new}
X_new <- lapply(1:3, function(i) matrix(rnorm(prod(p)), p[1], p[2]))
predict(fit, X_new, type = "response")
```

# Choosing how much to compress

`u[k]` is how many directions are kept in dimension `k`. Small values mean a
simpler, more stable model; larger values mean more flexibility and more
parameters (the model has `prod(u)` coefficients after compression, so
`u = c(2, 2)` costs four).

Omit `u` and an eigenvalue-ratio rule picks it:

```{r auto-u}
auto <- spgtr(X, y)
auto$u
```

If you would rather choose by predictive performance, fit a few and compare on
held-out subjects:

```{r choose-u}
train <- 1:150
test <- 151:200
sapply(1:3, function(k) {
  f <- spgtr(X[train], y[train], u = c(k, k))
  pred <- predict(f, X[test], type = "response")
  mean((y[test] - pred)^2) # Brier score, lower is better
})
```

# Selecting which parts of the array matter

Setting `lambda > 0` adds a penalty that removes entire rows and columns of the
array. `fit$selected` lists what survived in each dimension.

```{r sparse}
sparse <- spgtr(X, y, u = c(1, 1), lambda = 4)
sparse$nonzero # kept per dimension, out of 8 and 6
sparse$selected
```

Rather than guessing `lambda`, let cross-validation choose it. `spgtr_cv()`
fits the whole path in every fold, scores each value by held-out deviance, and
refits at the winner.

```{r cv}
set.seed(2)
cvfit <- spgtr_cv(X, y, u = c(1, 1), nfolds = 5, nlambda = 12)
cvfit$lambda_min
head(cvfit$cv)
cvfit$selected
```

```{r cvplot, fig.alt = "Cross-validated deviance against the sparsity level"}
plot(cvfit$cv$lambda, cvfit$cv$deviance, type = "b", log = "x",
     xlab = "lambda (sparsity)", ylab = "cross-validated deviance")
abline(v = cvfit$lambda_min, lty = 2)
```

The selected model uses only a few rows and columns, which is the interpretable
output most applications want: *these* regions of the array, and no others,
carry the signal.

```{r sparse-coef}
round(as.tensor(coef(cvfit))$as_array(), 2)
```

A caution: `lambda` selects **whole slices**, not individual cells. If row 3 is
kept, every cell in row 3 can be non-zero. That is the right notion when rows
and columns are meaningful units (brain regions, sensors, time points).

# Ordinary covariates

Age, sex, batch, and similar variables belong in `Z`. They enter the model
linearly, are never compressed, and are never penalized.

```{r covariates}
Z <- cbind(age = rnorm(n), male = rbinom(n, 1, 0.5))
fit_z <- spgtr(X, y, u = c(1, 1), Z = Z)
fit_z$gamma # one coefficient per column of Z
```

When a fit uses `Z`, prediction needs it too:

```{r covariates-predict}
head(predict(fit_z, X, Z, type = "response"))
```

# Outcomes other than yes/no

Pass any GLM family.

```{r families}
counts <- rpois(n, exp(1 + eta / 3))
fit_pois <- spgtr(X, counts, u = c(1, 1), family = poisson())
head(predict(fit_pois, type = "response"))

scores <- eta + rnorm(n, sd = 0.5)
fit_gauss <- spgtr(X, scores, u = c(1, 1), family = gaussian())
cor(predict(fit_gauss), scores)
```

For a continuous outcome with no sparsity you can also use `tepls()`, the
classical tensor envelope PLS estimator; `spgtr(..., family = gaussian(),
basis = "simpls")` reproduces it exactly.

```{r tepls-equivalence}
a <- tepls(X, scores, u = c(1, 1))
b <- spgtr(X, scores, u = c(1, 1), family = gaussian(), basis = "simpls")
max(abs(as.vector(a$coef$as_array()) - b$bvec))
```

# Honest evaluation

Everything `summary()` prints is in-sample and therefore optimistic. Split the
subjects, or read the cross-validated deviance from `spgtr_cv()`.

```{r holdout}
set.seed(3)
idx <- sample(n, 150)
f <- spgtr_cv(X[idx], y[idx], u = c(1, 1), nfolds = 5, nlambda = 8)
prob <- predict(f, X[-idx], type = "response")
yy <- y[-idx]

mean((prob > 0.5) == (yy == 1)) # accuracy
n1 <- sum(yy == 1); n0 <- sum(yy == 0)
(sum(rank(prob)[yy == 1]) - n1 * (n1 + 1) / 2) / (n1 * n0) # AUC
```

# What the algorithm does

For readers who want the mechanics. Write `X_i` for subject `i`'s array of
shape `p_1 x ... x p_m`, and `Z` for the ordinary covariates.

1. **Outcome enters through a working residual.** Fit the GLM of `y` on `Z`
   alone (the intercept only when `Z` is `NULL`) and take
   `r_i = y_i - mu_0i`. This is what makes the method work for any family: the
   rest of the algorithm only ever sees `r`.
2. **Second moments.** Center the predictors and form the mode-`k` marginal
   covariance `Sigma_k` (averaging over all other dimensions and subjects) and
   the cross-covariance array `C` between the centered predictor and `r`. The
   mode-`k` signal matrix is
   `U_k = C_(k) (kron_{j != k} Sigma_j^{-1}) C_(k)'`.
3. **Directions per dimension.** `W_k` (size `p_k x u_k`) is estimated from the
   pair `(U_k, Sigma_k)`. With `basis = "simpls"` it is the SIMPLS deflation of
   Zhang & Li (2017), a closed-form eigenvector recursion. With
   `basis = "envelope"` that basis is refined by minimizing the envelope
   objective `log|W' Sigma_k W| + log|W' (Sigma_k + U_k)^{-1} W|` over
   semi-orthogonal `W`.
4. **Sparsity.** `lambda > 0` adds `lambda * sum_i w_ki ||W_k[i, ]||_2`, an
   adaptively weighted group penalty on the *rows* of `W_k`. A zero row means
   slice `i` of dimension `k` cannot enter the model at all. The problem is
   solved by a proximal gradient method on the Stiefel manifold whose
   retraction is a right-multiplication, which is exactly what lets zeroed rows
   survive re-orthonormalization.
5. **Fit and translate back.** Reduce each subject to scores
   `T_i = X_i x_1 W_1' ... x_m W_m'`, fit the GLM of `y` on `(Z, vec(T))`, and
   map the latent coefficients `D` back with `B = D x_1 W_1 ... x_m W_m`. The
   returned `coef(fit)` is that `B`, stored as a `TTensor` (core `D`, factors
   `W`) so the structure is preserved; `as.tensor()` expands it.

References: Zhang & Li (2017, *Technometrics*) for the tensor PLS estimator,
Cook & Zhang (2016, *JCGS*) for envelope estimation, and Xiao, Liu & Yuan
(2021, *SIAM J. Optim.*) for the manifold proximal gradient solver.

# Performance notes

The two expensive steps -- the mode-wise covariances and the manifold solver --
are compiled kernels calling BLAS/LAPACK directly, and the score computation
reuses the package's compiled `ttm()`. Reference implementations in R are used
automatically if the package was built without compilation; the two paths agree
to numerical tolerance and are checked against each other in the test suite.

Practical guidance:

* A single `spgtr()` fit on a 60 × 60 predictor with 300 subjects takes about a
  tenth of a second; the full `spgtr_cv()` path (20 penalties, 5 folds) takes a
  second or two.
* Cost grows linearly in the number of subjects and in the total array size,
  and cubically in each individual dimension `p_k` (from the eigen problems),
  so a 200 × 200 dimension is far more expensive than four dimensions of 50.
* `basis = "simpls"` skips all iteration and is the fastest option when you do
  not need sparsity.
* `spgtr_cv()` computes covariances once per fold and warm-starts each penalty
  from the previous one, so a 20-point path costs far less than 20 separate
  fits.

# Limitations

* Sparsity is by slice, not by cell.
* All subjects must share one array shape; missing cells are not supported.
* In-sample statistics are optimistic; use held-out data.
* Standard errors and p-values are not provided. The model is selected from the
  data, so naive intervals would be wrong; use resampling if you need
  uncertainty.
* With a binary outcome and very strong signal the score-level GLM can separate
  perfectly, in which case `glm.fit()` warns and the coefficients grow. Reduce
  `u` or increase `lambda` if that happens.
