---
title: "Tensor PLS Regression with tepls()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tensor PLS Regression with tepls()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 3.2)
```

```{r setup}
library(tensory)
```

# The problem this solves

Each subject contributes a whole **array** — an image, a region-by-region
connectivity matrix, a sensor-by-time grid — and one or more **continuous**
outcomes to be predicted from it.

Flattening the array and calling `lm()` does not work. A 32 × 32 image is 1,024
predictors; with 200 subjects there is no unique least-squares solution, and
whatever a solver returns is fitting noise. Flattening also discards the grid:
it forgets that row 4 column 12 sits beside row 4 column 13.

`tepls()` keeps the grid. It compresses each *dimension* of the array
separately into a few informative directions, regresses the outcome on the
compressed array, then maps the answer back so the coefficient has the same
shape as the data. A 32 × 32 predictor reduced to 2 directions per mode leaves
4 numbers to estimate instead of 1,024.

**`tepls()` or `spgtr()`?**

|                        | `tepls()`                          | `spgtr()`                           |
|------------------------|------------------------------------|-------------------------------------|
| Outcome                | continuous; several at once is fine | any GLM family (binary, count, …)   |
| Extra covariates       | no                                 | yes, unpenalized                    |
| Selects whole slices   | no                                 | yes, via an L2,1 penalty            |
| Cost                   | one closed-form pass               | iterative                           |

For a continuous outcome and no covariates the two agree exactly when
`spgtr()` is given `family = gaussian()` and `basis = "simpls"`, which is
checked below. `tepls()` is the cheaper and more direct route in that setting;
`spgtr()` is the one for a non-Gaussian outcome, nuisance covariates, or slice
selection.

# How to lay out your data

Two forms are accepted, and they are interchangeable.

**A list, one array per subject.** Every element must have the same dimensions.

```{r}
set.seed(1)
p <- c(16, 12)
X <- lapply(1:150, function(i) matrix(rnorm(prod(p)), p[1], p[2]))
length(X)
dim(X[[1]])
```

**One big array with subjects in the *last* mode.** Data that already arrives
as a `16 x 12 x 150` block can be handed over directly, with no reshaping.

```{r}
Xbig <- array(unlist(X), dim = c(p, 150))
dim(Xbig)
```

The response is a numeric vector of length `n`, or an `n x r` matrix for
several outcomes per subject.

# Quick start

The outcome here depends on the array through a single rank-one pattern in the
top-left corner, plus noise.

```{r}
B_true <- outer(c(2, 1, rep(0, p[1] - 2)), c(1.5, rep(0, p[2] - 1)))
y <- vapply(X, function(xi) sum(B_true * xi), numeric(1)) + rnorm(150, sd = 0.5)

fit <- tepls(X, y)
fit
```

`u` was not supplied, so `tepls()` chose the number of directions per mode. The
estimated coefficient is an array of the same shape as a subject's data:

```{r}
B_hat <- as.tensor(coef(fit))$as_array()
dim(B_hat)
round(B_hat[1:4, 1:4], 2)
```

Only the corner is large, which is where the signal was planted:

```{r, fig.height = 3}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
image(t(B_true[p[1]:1, ]), main = "truth", axes = FALSE)
image(t(B_hat[p[1]:1, ]), main = "tepls()", axes = FALSE)
par(op)
```

Predictions come from `predict()`, with no `newdata` returning the fitted
values:

```{r}
pred <- predict(fit)
cor(pred, y)^2
```

# Choosing the number of directions

`u` is the one knob: how many directions to keep in each mode. `u = c(1, 1)`
keeps one row-pattern and one column-pattern (a rank-one coefficient);
`u = c(3, 2)` keeps three and two. Larger `u` means a more flexible fit and more
parameters — `prod(u)` of them.

Left at its default `NULL`, `tepls()` picks each mode's value from the largest
gap between consecutive eigenvalues of that mode's signal matrix — the same
rule `spgtr()` uses:

```{r}
fit$u
```

When you know the structure, say so:

```{r}
tepls(X, y, u = c(1, 1))$u
tepls(X, y, u = 2)$u # a scalar is recycled across modes
```

Otherwise cross-validate. There is no `tepls_cv()`; the fit is cheap enough
that an explicit loop over a grid does the job:

```{r}
cv_tepls <- function(X, y, u, nfolds = 5) {
  n <- length(X)
  fold <- sample(rep_len(seq_len(nfolds), n))
  err <- vapply(seq_len(nfolds), function(f) {
    tr <- which(fold != f)
    te <- which(fold == f)
    mean((predict(tepls(X[tr], y[tr], u = u), X[te]) - y[te])^2)
  }, numeric(1))
  mean(err)
}

set.seed(2)
grid <- list(c(1, 1), c(2, 1), c(2, 2), c(3, 3))
data.frame(
  u   = vapply(grid, paste, character(1), collapse = " x "),
  mse = vapply(grid, function(u) cv_tepls(X, y, u), numeric(1))
)
```

The prediction error is flat once `u` is large enough to hold the signal and
climbs as extra directions start fitting noise. Prefer the smallest `u` within
noise of the best.

# Several outcomes at once

Pass a matrix response and every column is fit jointly, sharing one set of
per-mode directions. The coefficient gains a trailing mode, one slice per
outcome.

```{r}
Y <- cbind(
  y,
  vapply(X, function(xi) sum((2 * B_true) * xi), numeric(1)) + rnorm(150, sd = 0.5)
)
mfit <- tepls(X, Y, u = c(1, 1))
mfit$r
dim(as.tensor(coef(mfit))$as_array())
dim(predict(mfit))
```

The two outcomes here are driven by the same image pattern, so fitting them
together estimates that pattern from twice the data.

# What is in the fit

```{r}
names(fit)
```

* `coef` — the coefficient `Tensor`, in the shape of the predictor (with a
  trailing response mode when there is more than one outcome). Also reachable
  as `coef(fit)`.
* `W` — one matrix per mode, `p_k x u[k]`, whose columns are the directions
  kept in that mode.
* `u`, `dims`, `r` — the shape of the problem.
* `intercept`, `Xbar` — the response mean and the predictor mean used for
  centering; `predict()` needs both.
* `fitted` — in-sample predictions.

The per-mode directions are interpretable on their own. The first mode's
direction should be concentrated on rows 1 and 2, and the second mode's on
column 1:

```{r}
round(fit$W[[1]][1:4, 1, drop = FALSE], 3)
round(fit$W[[2]][1:4, 1, drop = FALSE], 3)
```

The *latent scores* are the centered predictor projected onto those directions,
`prod(u)` numbers per subject — one apiece here. They serve for plotting or as
input to another model:

```{r}
scores <- vapply(X, function(xi) {
  as.vector(crossprod(fit$W[[1]], (xi - matrix(fit$Xbar, p[1], p[2]))) %*% fit$W[[2]])
}, numeric(prod(fit$u)))
str(scores)
```

# Out-of-sample prediction

In-sample `R^2` is optimistic. Hold data out.

```{r}
set.seed(3)
tr <- sample(150, 100)
te <- setdiff(1:150, tr)

f <- tepls(X[tr], y[tr], u = c(1, 1))
oos <- predict(f, X[te])

c(in_sample  = cor(predict(f), y[tr])^2,
  out_sample = cor(oos, y[te])^2)
```

`predict()` accepts new data in either input form, and checks that its
dimensions match the fit:

```{r}
identical(oos, predict(f, array(unlist(X[te]), dim = c(p, length(te)))))
try(predict(f, lapply(1:5, function(i) matrix(0, 3, 3))))
```

# How it works

`tepls()` implements Algorithm 4 of Zhang & Li (2017): each mode of the array
is compressed to a few directions chosen for their association with the
response, the response is regressed on the compressed predictor by least
squares, and the reduced coefficient is expanded back to the shape of the
original array. Nothing is iterative. No `prod(p) x prod(p)` matrix is
inverted, only the per-mode `p_k x p_k` ones, which is why `n` can be far
smaller than `prod(p)`. Zhang & Li (2017) give the derivation.

**Relation to the reference implementation.** The mode-`k` second-moment matrix
is identical to the `U U'` matrix in `TEReg::TensPLS_fit`. Two divergences are
deliberate: that package estimates each mode's basis with an envelope (`EnvMU`)
optimizer, whereas `tepls()` runs the SIMPLS deflation of Algorithm 4 exactly;
and the reduced regression here is plain least squares on the latent scores
(Algorithm 4, Step 6), which is invariant to the scale ambiguity of the
separable covariance, so no Kronecker scale needs pinning.

# Two identities worth knowing

**Keeping every direction is ordinary least squares.** With `u = p` nothing is
discarded, and the fit must coincide with `lm()` on the flattened predictor:

```{r}
set.seed(4)
ps <- c(3, 2)
Xs <- lapply(1:200, function(i) matrix(rnorm(prod(ps)), ps[1], ps[2]))
ys <- vapply(Xs, function(xi) sum(xi[1:2]), numeric(1)) + rnorm(200, sd = 0.2)

full <- tepls(Xs, ys, u = ps)
ols <- lm(ys ~ t(vapply(Xs, as.vector, numeric(prod(ps)))))
max(abs(as.vector(as.tensor(coef(full))$as_array()) - unname(coef(ols)[-1])))
```

**`spgtr()` with a Gaussian family and the SIMPLS basis is the same model.**

```{r}
a <- tepls(X, y, u = c(2, 2))
b <- spgtr(X, y, u = c(2, 2), family = gaussian(), basis = "simpls")
max(abs(as.vector(as.tensor(coef(a))$as_array()) - b$bvec))
```

# Reference

Zhang, X. and Li, L. (2017). Tensor envelope partial least-squares regression.
*Technometrics* **59**(4), 426–436.
