---
title: "Tensor PLS Regression with tepls()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tensor PLS Regression with tepls()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6,
                      fig.height = 3.2)
```

```{r setup}
library(tensory)
```

# The problem this solves

Each subject contributes a whole **array** — an image, a region-by-region
connectivity matrix, a sensor-by-time grid — and one or more **continuous**
outcomes. You want to predict the outcome from the array.

Flattening the array and calling `lm()` does not work. A 32 × 32 image is 1,024
predictors; with 200 subjects there is no unique least-squares solution, and
whatever a solver returns is fitting noise. Flattening also discards the grid:
it forgets that row 4 column 12 sits beside row 4 column 13.

`tepls()` keeps the grid. It compresses each *dimension* of the array
separately into a few informative directions, regresses the outcome on the
compressed array, and then maps the answer back so the coefficient has the same
shape as your data. A 32 × 32 predictor reduced to 2 directions per mode leaves
4 numbers to estimate instead of 1,024.

**`tepls()` or `spgtr()`?**

|                        | `tepls()`                          | `spgtr()`                           |
|------------------------|------------------------------------|-------------------------------------|
| Outcome                | continuous; several at once is fine | any GLM family (binary, count, …)   |
| Extra covariates       | no                                 | yes, unpenalized                    |
| Selects whole slices   | no                                 | yes, via an L2,1 penalty            |
| Cost                   | one closed-form pass               | iterative                           |

For a continuous outcome and no covariates the two agree exactly when
`spgtr()` is given `family = gaussian()` and `basis = "simpls"`; a check at the
end of this article confirms it. Reach for `tepls()` when that is your setting —
it is the cheaper, more direct route — and for `spgtr()` when you need a
non-Gaussian outcome, nuisance covariates, or slice selection.

# How to lay out your data

Two forms are accepted, and they are interchangeable.

**A list, one array per subject.** The most natural form. Every element must
have the same dimensions.

```{r}
set.seed(1)
p <- c(16, 12)
X <- lapply(1:150, function(i) matrix(rnorm(prod(p)), p[1], p[2]))
length(X)
dim(X[[1]])
```

**One big array with subjects in the *last* mode.** If your data already
arrives as a `16 x 12 x 150` block, hand it over directly — no reshaping.

```{r}
Xbig <- array(unlist(X), dim = c(p, 150))
dim(Xbig)
```

The response is a numeric vector of length `n`, or an `n x r` matrix if you
have several outcomes per subject.

# Quick start

We plant a signal in the top-left corner of the image: the outcome depends on
the array through a single rank-one pattern, plus noise.

```{r}
B_true <- outer(c(2, 1, rep(0, p[1] - 2)), c(1.5, rep(0, p[2] - 1)))
y <- vapply(X, function(xi) sum(B_true * xi), numeric(1)) + rnorm(150, sd = 0.5)

fit <- tepls(X, y)
fit
```

That is the whole call. `u` was not supplied, so `tepls()` chose the number of
directions per mode for you. The estimated coefficient is an array of the same
shape as a subject's data:

```{r}
B_hat <- as.tensor(coef(fit))$as_array()
dim(B_hat)
round(B_hat[1:4, 1:4], 2)
```

Only the corner is large, which is where we put the signal. A picture makes the
point faster than the numbers:

```{r, fig.height = 3}
op <- par(mfrow = c(1, 2), mar = c(2, 2, 2, 1))
image(t(B_true[p[1]:1, ]), main = "truth", axes = FALSE)
image(t(B_hat[p[1]:1, ]), main = "tepls()", axes = FALSE)
par(op)
```

Predictions come from `predict()`, with no `newdata` returning the fitted
values:

```{r}
pred <- predict(fit)
cor(pred, y)^2
```

# Choosing the number of directions

`u` is the one knob: how many directions to keep in each mode. `u = c(1, 1)`
keeps one row-pattern and one column-pattern (a rank-one coefficient);
`u = c(3, 2)` keeps three and two. Larger `u` means a more flexible fit and more
parameters — `prod(u)` of them.

Left at its default `NULL`, `tepls()` picks each mode's value from the largest
gap between consecutive eigenvalues of that mode's signal matrix — the same
rule `spgtr()` uses. It is fast and usually sensible:

```{r}
fit$u
```

When you know the structure, say so:

```{r}
tepls(X, y, u = c(1, 1))$u
tepls(X, y, u = 2)$u # a scalar is recycled across modes
```

When you do not, cross-validate. There is no `tepls_cv()`; the fit is cheap
enough that an explicit loop is clearer than a wrapper:

```{r}
cv_tepls <- function(X, y, u, nfolds = 5) {
  n <- length(X)
  fold <- sample(rep_len(seq_len(nfolds), n))
  err <- vapply(seq_len(nfolds), function(f) {
    tr <- which(fold != f)
    te <- which(fold == f)
    mean((predict(tepls(X[tr], y[tr], u = u), X[te]) - y[te])^2)
  }, numeric(1))
  mean(err)
}

set.seed(2)
grid <- list(c(1, 1), c(2, 1), c(2, 2), c(3, 3))
data.frame(
  u   = vapply(grid, paste, character(1), collapse = " x "),
  mse = vapply(grid, function(u) cv_tepls(X, y, u), numeric(1))
)
```

The prediction error is flat once `u` is large enough to hold the signal and
climbs as extra directions start fitting noise. Prefer the smallest `u` within
noise of the best.

# Several outcomes at once

Pass a matrix response and every column is fit jointly, sharing one set of
per-mode directions. The coefficient gains a trailing mode, one slice per
outcome.

```{r}
Y <- cbind(
  y,
  vapply(X, function(xi) sum((2 * B_true) * xi), numeric(1)) + rnorm(150, sd = 0.5)
)
mfit <- tepls(X, Y, u = c(1, 1))
mfit$r
dim(as.tensor(coef(mfit))$as_array())
dim(predict(mfit))
```

Sharing the directions is the point: the two outcomes here are driven by the
same image pattern, and fitting them together estimates that pattern from twice
the data.

# What is in the fit

```{r}
names(fit)
```

* `coef` — the coefficient `Tensor`, in the shape of your predictor (with a
  trailing response mode when there is more than one outcome). Also reachable
  as `coef(fit)`.
* `W` — one matrix per mode, `p_k x u[k]`, whose columns are the directions
  kept in that mode. These are what the method learns.
* `u`, `dims`, `r` — the shape of the problem.
* `intercept`, `Xbar` — the response mean and the predictor mean used for
  centering; `predict()` needs both.
* `fitted` — in-sample predictions.

The per-mode directions are interpretable on their own. Here the first mode's
direction should be concentrated on rows 1 and 2, and the second mode's on
column 1:

```{r}
round(fit$W[[1]][1:4, 1, drop = FALSE], 3)
round(fit$W[[2]][1:4, 1, drop = FALSE], 3)
```

Subjects' compressed coordinates — the *latent scores* — are the centered
predictor projected onto those directions — `prod(u)` numbers per subject, so
just one apiece here. They are useful for plotting or as input to another
model:

```{r}
scores <- vapply(X, function(xi) {
  as.vector(crossprod(fit$W[[1]], (xi - matrix(fit$Xbar, p[1], p[2]))) %*% fit$W[[2]])
}, numeric(prod(fit$u)))
str(scores)
```

# Evaluating honestly

In-sample `R^2` is optimistic. Hold data out.

```{r}
set.seed(3)
tr <- sample(150, 100)
te <- setdiff(1:150, tr)

f <- tepls(X[tr], y[tr], u = c(1, 1))
oos <- predict(f, X[te])

c(in_sample  = cor(predict(f), y[tr])^2,
  out_sample = cor(oos, y[te])^2)
```

`predict()` accepts new data in either input form, and checks that its
dimensions match the fit:

```{r}
identical(oos, predict(f, array(unlist(X[te]), dim = c(p, length(te)))))
try(predict(f, lapply(1:5, function(i) matrix(0, 3, 3))))
```

# How it works

`tepls()` implements Algorithm 4 of Zhang & Li (2017).

1. **Center** the predictor and response.
2. **Marginal covariances.** For each mode `k`, `Sigma_k` averages
   `X_(k) X_(k)'` over subjects — how variable the array is along that mode.
   (This step runs in C++ when the compiled kernel is available.)
3. **Cross-covariance.** `C` is the array of covariances between each predictor
   cell and the response — where the signal is.
4. **Per-mode signal matrix.** `M_k = C_(k) (kron_{j != k} Sigma_j^-1) C_(k)'`
   standardizes `C` along every mode but `k`, so `M_k` measures signal in mode
   `k` after accounting for the others.
5. **SIMPLS deflation.** The leading eigenvector of `M_k` is the first
   direction; project it out (in the `Sigma_k` metric) and repeat, `u[k]` times.
   This gives `W_k`.
6. **Reduced regression.** Project every subject onto `W_1 kron ... kron W_m`
   and regress the response on the resulting `prod(u)` scores by least squares.
7. **Map back.** `B = W (reduced coefficient)`, reshaped to the predictor's
   shape (their Lemma 2).

Nothing here is iterative, and nothing inverts a `prod(p) x prod(p)` matrix —
only the per-mode `p_k x p_k` ones. That is why `n` can be far smaller than
`prod(p)`.

**Relation to the reference implementation.** The mode-`k` second-moment matrix
built in step 4 is identical to the `U U'` matrix in `TEReg::TensPLS_fit`. Two
deliberate divergences: that package estimates each mode's basis with an
envelope (`EnvMU`) optimizer, whereas `tepls()` runs the SIMPLS deflation of
Algorithm 4 exactly; and the reduced regression here is plain least squares on
the latent scores (Step 6), which is invariant to the scale ambiguity of the
separable covariance, so no Kronecker scale needs pinning.

# Two identities worth knowing

**Keeping every direction is ordinary least squares.** With `u = p` nothing is
discarded, and the fit must coincide with `lm()` on the flattened predictor:

```{r}
set.seed(4)
ps <- c(3, 2)
Xs <- lapply(1:200, function(i) matrix(rnorm(prod(ps)), ps[1], ps[2]))
ys <- vapply(Xs, function(xi) sum(xi[1:2]), numeric(1)) + rnorm(200, sd = 0.2)

full <- tepls(Xs, ys, u = ps)
ols <- lm(ys ~ t(vapply(Xs, as.vector, numeric(prod(ps)))))
max(abs(as.vector(as.tensor(coef(full))$as_array()) - unname(coef(ols)[-1])))
```

**`spgtr()` with a Gaussian family and the SIMPLS basis is the same model.**

```{r}
a <- tepls(X, y, u = c(2, 2))
b <- spgtr(X, y, u = c(2, 2), family = gaussian(), basis = "simpls")
max(abs(as.vector(as.tensor(coef(a))$as_array()) - b$bvec))
```

If either identity ever breaks, something is wrong with the centering, the
score map, or the reconstruction — both are in the test suite for that reason.

# Limits

* **Continuous outcomes only.** Binary or count outcomes belong in `spgtr()`.
* **No covariates.** Age, sex, batch and the like cannot be entered
  unpenalized; `spgtr()`'s `Z` argument does that.
* **No sparsity.** Every row and column of the array contributes. When you need
  "which slices matter", use `spgtr()` with `lambda > 0`.
* **No standard errors.** The coefficient is a point estimate. Bootstrap the
  subjects if you need uncertainty.
* **`u` is not free.** Choose it by cross-validation when the eigenvalue-gap
  default is not obviously right — with `prod(u)` parameters, a too-large `u`
  overfits like any other model.

# Reference

Zhang, X. and Li, L. (2017). Tensor envelope partial least-squares regression.
*Technometrics* **59**(4), 426–436.
